Computationally Efficient Simulation of Queues: The \proglang{R} Package \pkg{queuecomputer}

### 2018-11-06 main.tex
Minor details.

The package is modular and integrates well with the popular \proglang{R} package \pkg{dplyr}. Complex queueing networks with tandem, parallel and fork/join topologies can easily be built with these two packages together. We show how to use this package with two examples: a call-centre and an airport terminal. 
}
\Keywords{queues, queueing theory, discrete event simulation, 
operations research, approximate Bayesian computation, \proglang{R}}
\Plainkeywords{queues, queueing theory, discrete event simulation, 
operations research, approximate Bayesian computation, R} %% without formatting
%% at least one keyword must be supplied

@@

% This is a useful model for many real-world applications where the customer/server analogy applies such as: \citet{koole2002queueing} (call centres); \citet{lakshmi2013application} (hospitals); \cite{cao2003web} (web servers) and \citet{papadopoulos1996queueing} (manufacturing systems). For a detailed introduction to queueing theory see \citet{medhi2002stochastic,bhat2015introduction}.

The queues we encounter in our everyday experience, where customers wait in line to be served by a server, is a useful analogy for many other processes. We say analogy because the word customers could represent: MapReduce jobs \citep{lin_joint_2013}; patients in a hospital \citep{takagi_queueing_2016}; items in a manufacturing system \citep{dallery_manufacturing_1992}; calls to a call center \citep{gans_telephone_2003}; shipping containers in a seaport \citep{kozan_comparison_1997} or even cognitive tasks \citep{cao_queueing_2013}. Similarly server could represent: a compute cluster; medical staff; machinery or a customer service representative at a call centre. Queueing systems can also be networked together to form queueing networks. We can use queueing networks to build models of processes such as provision of internet services \citep{sutton_bayesian_2011}, passenger facilitation at international airports \citep{wu_review_2013} and emergency evacuations \citep{van_woensel_modeling_2007}. Clearly queueing systems and queueing networks are useful for understanding important real-world systems.

@@

In Section \ref{sec:queueing} we review the literature on queueing theory and develop notation used throughout this paper. In Section \ref{sec:QDC} we present the QDC algorithm and compare it to DES. We demonstrate usage of the package in Section \ref{sec:usage}. Details of implementation and usage are discussed in Section \ref{sec:Implementation}. The package is validated in Section \ref{sec:Validation} by replicating results from DES packages \pkg{simpy} and \pkg{simmer}. We compare computed performance measures from the output of a \pkg{queuecomputer} simulation to theoretical results for $M/M/2$ queueing systems. We benchmark the package in Section \ref{sec:Benchmark} and compare computation time with \pkg{simpy} and \pkg{simmer}. Examples in Section \ref{sec:Examples} are used to demonstrate how the package can be used to simulate a call centre and an international airport terminal.

@@

The characteristics of a queueing system is expressed with the notation of \citet{kendall1953stochastic}. This notation has since been extended to six characteristics:
\begin{itemize}
\item $f_{\mathbf{\delta}}$, inter-arrival distribution;
\item $f_{\mathbf{s}}$, service distribution; 
\item $K$, number of servers $\in \mathbb{N}$;
\item $C$, capacity of system $\in \mathbb{N}$; 
\item $n$, customer population $\in \mathbb{N}$; and
\item $R$, service discipline
\end{itemize}

@@

\begin{align}
\Prob(0) &= \left[\frac{ (K \rho)^K}{K! (1 - \rho)} + 1 + \sum_{i=1}^{K-1} \frac{(K \rho)^i}{i!}   \right]^{-1} \\
\Prob(N) &= \begin{cases} 
\Prob(0) \frac{(K \rho)^n }{N!} & \quad  N \leq K\\
\Prob(0) \frac{(K \rho)^n }{K! K^{N-K}} & \text{otherwise}
\end{cases} \label{eq:pn}
\end{align}

@@

\begin{align}
\E(N) = K\rho + \frac{\rho (K\rho)^K P(0)}{K!(1-\rho)^2},
\end{align}

@@

\begin{align}
\E(\mathbf{w}) = \frac{(K \rho)^K P(0)}{K! K \mu(1-\rho)^2}.
\end{align}

If the parameters of $f_\delta$ and $f_{\mathbf{s}}$ are uncertain then we must turn to predictive distributions for estimates of performance measures. Predictive distributions of performance measures using Bayesian posterior distributions are derived by \citet{armero1994bayesian, armero1999dealing}.

@@

Most models of queueing systems assume time-invariant inter-arrival and service processes. In practice many real-world queues have inter-arrival processes which are strongly time-dependent, such as: call-centres \citep{weinberg2007bayesian, brown2005statistical}, airport runways \citep{koopman1972air} and hospitals \citep{brahimi_queueing_1991}. In the case of the $M/M/1$ queue we can adapt the notation to $M(t)/M(t)/1$ to represent exponential  processes where parameters $\lambda(t)$ and $\mu(t)$ change with time. Such queueing systems are referred to as dynamic queueing systems.

@@

QDC can be considered as a multi-server extension to an algorithm presented by \citet{lindley_theory_1952}. For a single server queueing system, the departure time of the $i$th customer is: $d_i = \max{(a_i, d_{i-1})} + s_i$, since the customer either waits for a server or the server waits for a customer. The algorithm (not the paper) was, surprisingly, not extended to multi-server systems until \citet{krivulin_recursive_1994}. However with each new customer $i$ the algorithm must search a growing $i+1$ length vector. This algorithm therefore scales poorly, with computational complexity $O(n^2)$, where $n$ is number of customers. \citet{kin_generalized_2010} adapted the original algorithm of \citet{kiefer_theory_1955} to an $O(nK)$ algorithm for multiserver tandem queues with blocking, that is $G/G/K/C$ queueing systems where $C$ is the maximum capacity number of customers in the queueing systems.

@@

then we need consider a change in state over at most 1 knot location. This step function is determined input by the user. Like the arrival and service times $\mathbf{(a,s)}$ it is changeable by the user before the simulation but not during the simulation.

@@

The DES packages \pkg{simmer} and \pkg{simpy} are not built to allow users to input $(\mathbf{a,s})$ directly. Rather, the user supplies parameters for $f_{\delta}$ and $f_{\mathbf{s}}$ so that inter-arrival and service times can be sampled at each step when needed. To allow \pkg{simmer} and \pkg{simpy} to accept presampled input $(\mathbf{a,s})$ we use generator functions in place of the usual \code{rexp(rate)} or \code{random.expovariate(rate)} calls in \proglang{R} and \proglang{Python} respectively, details of this work can be found in the supplementary material. We create an interface to \pkg{simmer} so that it can be called in the same way as \pkg{queuecomputer}.

@@

\subsection{Call centre} \label{ssec:callcentre}

We demonstrate \pkg{queuecomputer} by simulating a call centre. The arrival time for each customer is the time that they called, the service time is how long it takes for their problem to be resolved once they reach an available customer service representative. Let's assume that the customers arrive by a homogeneous Poisson process over the course of the day.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width = 0.8\textwidth]{figures/Larger_airport.pdf}
\caption{Diagram of larger airport scenario, there are 120 flights in total and two multi-server queueing systems operate in parallel. Passengers are preassigned to travel through either the ``manual" or ``smart gate" route through immigration. The passengers and bags are ``forked'' when each aircraft arrives are are ``joined'' at the baggage hall. }
\label{fig:larger_airport}
\end{figure}

@@

The \proglang{R} package \pkg{queuecomputer} implements QDC. It can be used to simulate any queueing systems or tandem network of queueing systems of general form $G(t)/G(t)/K(t)/\infty/n/FCFS$. Fast algorithms for multi-server queueing systems have been proposed in the past \citep{krivulin_recursive_1994,sutton_inference_2010,kin_generalized_2010}. These algorithms have generated little notice, even in the cases where their computational efficiency is demonstrated \citep{kin_generalized_2010}. QDC is conceptually simpler, more efficient memory-wise and modular.

@@

This work is supported by the Australian Research Council Centre of Excellence for Mathematical and Statistical Frontiers (ACEMS). This work was funded through the Australian Research Council (ARC) linkage grant ``Improving the Productivity and Efficiency of Australian Airports” (LP140100282).

>>>>

The package is modular and integrates well with the popular \proglang{R} package \pkg{dplyr}. Complex queueing networks with tandem, parallel and fork/join topologies can easily be built with these two packages together. We show how to use this package with two examples: a call-center and an airport terminal. 
}
\Keywords{queues, queueing theory, discrete event simulation, 
operations research, approximate Bayesian computation, \proglang{R}}
\Plainkeywords{queues, queueing theory, discrete event simulation, 
operations research, approximate Bayesian computation, R} %% without formatting
%% at least one keyword must be supplied

@@

% This is a useful model for many real-world applications where the customer/server analogy applies such as: \citet{koole2002queueing} (call centers); \citet{lakshmi2013application} (hospitals); \cite{cao2003web} (web servers) and \citet{papadopoulos1996queueing} (manufacturing systems). For a detailed introduction to queueing theory see \citet{medhi2002stochastic,bhat2015introduction}.

The queues we encounter in our everyday experience, where customers wait in line to be served by a server, are a useful analogy for many other processes. We say analogy because the word customers could represent: MapReduce jobs \citep{lin_joint_2013}; patients in a hospital \citep{takagi_queueing_2016}; items in a manufacturing system \citep{dallery_manufacturing_1992}; calls to a call center \citep{gans_telephone_2003}; shipping containers in a seaport \citep{kozan_comparison_1997} or even cognitive tasks \citep{cao_queueing_2013}. Similarly server could represent: a compute cluster; medical staff; machinery or a customer service representative at a call center. Queueing systems can also be networked together to form queueing networks. We can use queueing networks to build models of processes such as provision of internet services \citep{sutton_bayesian_2011}, passenger facilitation at international airports \citep{wu_review_2013} and emergency evacuations \citep{van_woensel_modeling_2007}. Clearly queueing systems and queueing networks are useful for understanding important real-world systems.

@@

In Section \ref{sec:queueing} we review the literature on queueing theory and develop notation used throughout this paper. In Section \ref{sec:QDC} we present the QDC algorithm and compare it to DES. We demonstrate usage of the package in Section \ref{sec:usage}. Details of implementation and usage are discussed in Section \ref{sec:Implementation}. The package is validated in Section \ref{sec:Validation} by replicating results from DES packages \pkg{simpy} and \pkg{simmer}. We compare computed performance measures from the output of a \pkg{queuecomputer} simulation to theoretical results for $M/M/2$ queueing systems. We benchmark the package in Section \ref{sec:Benchmark} and compare computation time with \pkg{simpy} and \pkg{simmer}. Examples in Section \ref{sec:Examples} are used to demonstrate how the package can be used to simulate a call center and an international airport terminal.

@@

The characteristics of a queueing system are expressed with the notation of \citet{kendall1953stochastic}. This notation has since been extended to six characteristics:
\begin{itemize}
\item $f_{\mathbf{\delta}}$, inter-arrival distribution;
\item $f_{\mathbf{s}}$, service distribution; 
\item $K$, number of servers $\in \mathbb{N}$;
\item $C$, capacity of system $\in \mathbb{N}$; 
\item $n$, customer population $\in \mathbb{N}$; and
\item $R$, service discipline
\end{itemize}

@@

\begin{align}
\Prob(0) &= \left[\frac{ (K \rho)^K}{K! (1 - \rho)} + 1 + \sum_{i=1}^{K-1} \frac{(K \rho)^i}{i!}   \right]^{-1} \nonumber \\
\Prob(N) &= \begin{cases} 
\Prob(0) \frac{(K \rho)^n }{N!} & \quad  N \leq K\\
\Prob(0) \frac{(K \rho)^n }{K! K^{N-K}} & \text{otherwise} 
\end{cases} \label{eq:pn}
\end{align}

@@

\begin{align}
\E(N) = K\rho + \frac{\rho (K\rho)^K P(0)}{K!(1-\rho)^2} \label{eq:en},
\end{align}

@@

\begin{align}
\E(\mathbf{w}) = \frac{(K \rho)^K P(0)}{K! K \mu(1-\rho)^2} \label{eq:ew}.
\end{align}

If the parameters of $f_\delta$ and $f_{\mathbf{s}}$ are uncertain then we must turn to predictive distributions for estimates of performance measures, where are computed analytically for M/M/K queues (Equations \ref{eq:en} and \ref{eq:ew}). Predictive distributions of performance measures using Bayesian posterior distributions are derived by \citet{armero1994bayesian, armero1999dealing}.

@@

Most models of queueing systems assume time-invariant inter-arrival and service processes. In practice many real-world queues have inter-arrival processes which are strongly time-dependent, such as: call-centers \citep{weinberg2007bayesian, brown2005statistical}, airport runways \citep{koopman1972air} and hospitals \citep{brahimi_queueing_1991}. In the case of the $M/M/1$ queue we can adapt the notation to $M(t)/M(t)/1$ to represent exponential  processes where parameters $\lambda(t)$ and $\mu(t)$ change with time. Such queueing systems are referred to as dynamic queueing systems.

@@

QDC can be considered as a multiserver extension to an algorithm presented by \citet{lindley_theory_1952}. For a single server queueing system, the departure time of the $i$th customer is: $d_i = \max{(a_i, d_{i-1})} + s_i$, since the customer either waits for a server or the server waits for a customer. The algorithm (not the paper) was, surprisingly, not extended to multiserver systems until \citet{krivulin_recursive_1994}. However with each new customer $i$ the algorithm must search a growing $i+1$ length vector. This algorithm therefore scales poorly, with computational complexity $O(n^2)$, where $n$ is number of customers. \citet{kin_generalized_2010} adapted the original algorithm of \citet{kiefer_theory_1955} to an $O(nK)$ algorithm for multiserver tandem queues with blocking, that is $G/G/K/C$ queueing systems where $C$ is the maximum capacity number of customers in the queueing systems.

@@

then we need to consider a change in state over at most 1 knot location. This step function is determined input by the user. Like the arrival and service times $\mathbf{(a,s)}$ it is changeable by the user before the simulation but not during the simulation.

@@

The DES packages \pkg{simmer} and \pkg{simpy} are not built to allow users to input $(\mathbf{a,s})$ directly. Rather, the user supplies parameters for $f_{\delta}$ and $f_{\mathbf{s}}$ so that inter-arrival and service times can be sampled at each step when needed. To allow \pkg{simmer} and \pkg{simpy} to accept presampled input $(\mathbf{a,s})$ we use generator functions instead of \code{rexp(rate)} or \code{random.expovariate(rate)} calls in \proglang{R} and \proglang{Python} respectively, details of this work can be found in the supplementary material. We create an interface to \pkg{simmer} so that it can be called in the same way as \pkg{queuecomputer}.

@@

\subsection{Call center} \label{ssec:callcenter}

We demonstrate \pkg{queuecomputer} by simulating a call center. The arrival time for each customer is the time that they called, the service time is how long it takes for their problem to be resolved once they reach an available customer service representative. Let's assume that the customers arrive by a homogeneous Poisson process over the course of the day.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width = 0.8\textwidth]{figures/Larger_airport.pdf}
\caption{Diagram of larger airport scenario, there are 120 flights in total and two multiserver queueing systems operate in parallel. Passengers are preassigned to travel through either the ``manual" or ``smart gate" route through immigration. The passengers and bags are ``forked'' when each aircraft arrives are are ``joined'' at the baggage hall. }
\label{fig:larger_airport}
\end{figure}

@@

The \proglang{R} package \pkg{queuecomputer} implements QDC. It can be used to simulate any queueing systems or tandem network of queueing systems of general form $G(t)/G(t)/K(t)/\infty/n/FCFS$. Fast algorithms for multiserver queueing systems have been proposed in the past \citep{krivulin_recursive_1994,sutton_inference_2010,kin_generalized_2010}. These algorithms have generated little notice, even in the cases where their computational efficiency is demonstrated \citep{kin_generalized_2010}. QDC is conceptually simpler, more efficient memory-wise and modular.

@@

This work is supported by the Australian Research Council center of Excellence for Mathematical and Statistical Frontiers (ACEMS). This work was funded through the Australian Research Council (ARC) linkage grant ``Improving the Productivity and Efficiency of Australian Airports” (LP140100282).
